### 2015-02-27 icfp15-servo/abstract.tex
Fix nits

%!TEX root = paper.tex
%
\begin{abstract}
All modern web browsers --- Internet Explorer, Firefox, Chrome, Opera, and Webkit ---
have a core rendering engine written in \Cplusplus.
This language choice was made because it affords the systems programmer complete control
of the underlying hardware features and memory in use plus provides a transparent
compilation model.

Servo is a project started at Mozilla Research to build a new web browser engine that
preserves the capabilities of these other browser engines but also both takes more
advantage of the recent trends in parallel hardware and is more memory-safe.
We use a new language, Rust, that provides us a similar level of control of the underlying
system but which builds on many concepts familiar to the functional programming community
to form a novelty - a useful, safe systems programming language.

>>>>

%!TEX root = paper.tex
%
\begin{abstract}
All modern web browsers --- Internet Explorer, Firefox, Chrome, Opera, and Safari ---
have a core rendering engine written in \Cplusplus{}.
This language choice was made because it affords the systems programmer complete control
of the underlying hardware features and memory in use, and it provides a transparent
compilation model.

Servo is a project started at Mozilla Research to build a new web browser engine that
preserves the capabilities of these other browser engines but also both takes 
advantage of the recent trends in parallel hardware and is more memory-safe.
We use a new language, Rust, that provides us a similar level of control of the underlying
system to \Cplusplus{} but which builds on many concepts familiar to the functional programming
community, forming a novelty --- a useful, safe systems programming language.

### 2015-02-27 icfp15-servo/intro.tex
Fix nits

%!TEX root = paper.tex
\section{Introduction}
\label{sec:intro}
The heart of a modern web browser is the browser engine, which is the code responsible
for loading, processing, evaluating, and rendering web content.
There are three current browser engine families:
\begin{enumerate}
\item Trident/Spartan, the engine in Internet Explorer~\cite{IE}
\item Webkit\cite{WEBKIT}/Blink, the engine in Safari~\cite{SAFARI}, Chrome~\cite{CHROME}, and Opera~\cite{OPERA}
\item Gecko, the engine in Firefox~\cite{FIREFOX}
\end{enumerate}
All of these engines have at their core many millions of lines of \Cplusplus{} code.
While the use of \Cplusplus{} has enabled all of these browsers to achieve excellent sequential
performance on a single web page, on mobile devices with lower processor speed but many
more processors, these browsers do not provide the same level of interactivity that they
do on desktop processors~\cite{parallelizing-web-pages,ZOOMM}.
Further, in an informal inspection of the critical security bugs in Gecko, we determined that
roughly 50\% of the bugs are use after free, out of range access, or related to integer
overflow.
The other 50\% is split between errors in tracing values from the JavaScript heap in the
\Cplusplus{} code and errors related to dynamically compiled code.

Servo~\cite{SERVO} is a new web browser engine designed to address the major environment and 
architectural changes over the last decade.
The goal of the Servo project is to produce a browser that enables new applications to be authored
against the web platform that run with more safety, better performance, and better power usage
than in current browsers.
To address memory-related safety issues, we are using a new systems programming language,
Rust~\cite{RUST}.
For parallelism and power, we scale across a wide variety of hardware by building either data-
or task-parallelism, as appropriate, into each part of the web platform.
Additionally, we are improving concurrency by reducing the simultaneous access to data
structures and using a message-passing architecture between components such as the
JavaScript engine and the rendering engine that paints graphics to the screen.
Servo is currently over 300k lines of code and implements enough of the web to render and
process many pages, though it is still a far cry from the over 7 million lines of code in
the Mozilla Firefox browser and its associated libraries.
However, we believe that we have implemented enough of the web platform to provide an
early report on the successes, failures, and open problems remaining in Servo, from the
point of view of programming languages and runtime research.
In this experience report, we discuss the design and architecture of a modern web 
browser engine, show how modern programming language techniques --- many of which
originated in the functional programming community --- address these design 
constraints, and also touch on ongoing challenges and areas of research where we
would welcome additional community input.

>>>>

%!TEX root = paper.tex
\section{Introduction}
\label{sec:intro}
The heart of a modern web browser is the browser engine, which is the code responsible
for loading, processing, evaluating, and rendering web content.
There are three major browser engine families:
\begin{enumerate}
\item Trident/Spartan, the engine in Internet Explorer~\cite{IE}
\item Webkit\cite{WEBKIT}/Blink, the engine in Safari~\cite{SAFARI}, Chrome~\cite{CHROME}, and Opera~\cite{OPERA}
\item Gecko, the engine in Firefox~\cite{FIREFOX}
\end{enumerate}
All of these engines have at their core many millions of lines of \Cplusplus{} code.
While the use of \Cplusplus{} has enabled all of these browsers to achieve excellent sequential
performance on a single web page, on mobile devices with lower processor speed but many
more processors, these browsers do not provide the same level of interactivity that they
do on desktop processors~\cite{parallelizing-web-pages,ZOOMM}.
Further, in an informal inspection of the critical security bugs in Gecko, we determined that
roughly 50\% of the bugs are use after free, out of range access, or related to integer
overflow.
The other 50\% are split between errors in tracing values from the JavaScript heap in the
\Cplusplus{} code and errors related to dynamically compiled code.

Servo~\cite{SERVO} is a new web browser engine designed to address the major environment and 
architectural changes over the last decade.
The goal of the Servo project is to produce a browser that enables new applications to be authored
against the web platform that run with more safety, better performance, and better power usage
than in current browsers.
To address memory-related safety issues, we are using a new systems programming language,
Rust~\cite{RUST}.
For parallelism and power, we scale across a wide variety of hardware by building either data-
or task-parallelism, as appropriate, into each part of the web platform.
Additionally, we are improving concurrency by reducing the simultaneous access to data
structures and using a message-passing architecture between components such as the
JavaScript engine and the rendering engine that paints graphics to the screen.
Servo is currently over 400k lines of Rust code and implements enough of the web to render and
process many pages, though it is still a far cry from the over 7 million lines of code in
the Mozilla Firefox browser and its associated libraries.
However, we believe that we have implemented enough of the web platform to provide an
early report on the successes, failures, and open problems remaining in Servo, from the
point of view of programming languages and runtime research.
In this experience report, we discuss the design and architecture of a modern web 
browser engine, show how modern programming language techniques --- many of which
originated in the functional programming community --- address these design 
constraints, and also touch on ongoing challenges and areas of research where we
would welcome additional community input.

### 2015-02-27 icfp15-servo/servo.tex
Fix nits

\section{Servo}
\label{sec:servo}
In this section, we cover specific areas of Servo's design or implementation that make use of Rust
and the impacts and limitations of these features.
The most important test for a browser, though, is performance --- Servo must be at least as fast
as other browsers at similar tasks to succeed, even if it provides additional memory safety.
\tblref{servo-perf} shows a preliminary breakdown of the performance in the layout stage (described
in \secref{sec:browsers}) of rendering several web sites, taken on a modern Macbook Pro.
\begin{table}
  \begin{center}
    \begin{tabular}{r || r | r | r}
      Site & Gecko & Servo 1 thread & Servo 4 threads \\
      \hline
      Reddit & 250 & 100 & 55  \\
      CNN & 105 & 50 & 35 \\
    \end{tabular}%
  \end{center}%
  \caption{Performance of Servo against Mozilla's Gecko rendering engine on the layout portion of some common sites.
  Times are in milliseconds.}
  \label{servo-perf}
\end{table}

@@

\subsection{Compilation strategy}
Many statically typed implementations of polymorphic languages such as Standard ML of New Jersey~\cite{SMLNJ} and
\ocaml{}~\cite{ocaml-manual-3.0} have used a compilation strategy that optimizes representations of data types when
polymorphic code is monomorphically used, but defaults to a less efficient style otherwise, in order to share
code~\cite{ocaml-repr}.
This strategy reduces code size, but leads to unpredictable performance and code, as changes to a codebase that
either add a new instantiation of a polymorphic function at a given type or, in a modular compilation setting, that
expose a polymorphic function externally, can change the performance of code that is not local to the change being
made.
Monomorphization, as in MLton~\cite{mlton-compiler}, instead instantiates each polymorphic code block at each of the types
it is applied against, providing predictable output code to developers at the cost of some code duplication.
This approach, combined with whole-program compilation~\cite{weeks:whole-program-mlton}, has a negative impact on
compilation speed but has resulted in Rust code that easily matches the sequential speed of its \Cplusplus{} analog
without requiring the Servo developers to become compiler experts.

@@

Additionally, Rust's ownership model assumes that there is a single owner for each piece of data.
However, many data structures do not follow that model, in order to provide asymptotic
improvements in access.
For example, a doubly-linked list contains a back pointer to the previous element to aid
in traversals in the opposite direction.
Many optimized hashtable implementations also have both hash-based access to items
and a linked list of all of the keys or values.
In Servo, we have had to use unsafe code to implement data structures with this form,
though we are typically able to provide a safe interface to users.

@@

There are two limitations in the language interoperability that pose challenges for Servo today.
First, Rust cannot currently expose varargs-style functions to C code.
Second, Rust cannot compile against \Cplusplus{} code.
In both cases, Servo uses C wrapper code to call into the code that Rust cannot directly
reach.
While this approach is not a large problem for varargs-style functions, it defeats many of the
places where the \Cplusplus{} code has been crafted to ensure the code is inlined into the caller,
which results in degraded performance.

@@

Compiler plugins can also provide ``lint'' checks\footnote{\url{http://doc.rust-lang.org/book/plugins.html#lint-plugins}} that use the same infrastructure as the compiler's built-in warnings. This allows project-specific safety or style checks to integrate deeply with the compiler. Lint plugins traverse a typechecked AST, and they can be enabled/disabled within any lexical scope, the same way as built-in warnings.

>>>>

\section{Servo}
\label{sec:servo}
A crucial test of Servo is performance --- Servo must be at least as fast
as other browsers at similar tasks to succeed, even if it provides additional memory safety.
\tblref{servo-perf} shows a preliminary comparison of the performance of the layout stage (described
in \secref{sec:browsers}) of rendering several web sites in Mozilla Firefox's Gecko engine compared
to Servo, taken on a modern Macbook Pro.
\begin{table}
  \begin{center}
    \begin{tabular}{r || r | r | r}
      Site & Gecko & Servo 1 thread & Servo 4 threads \\
      \hline
      Reddit & 250 & 100 & 55  \\
      CNN & 105 & 50 & 35 \\
    \end{tabular}%
  \end{center}%
  \caption{Performance of Servo against Mozilla's Gecko rendering engine on the layout portion of some common sites.
  Times are in milliseconds, where lower numbers are better.}
  \label{servo-perf}
\end{table}
In the remainder of this section, we cover specific areas of Servo's design or implementation that make use of
Rust and the impacts and limitations of these features.

@@

\subsection{Compilation strategy}
Many statically typed implementations of polymorphic languages such as Standard ML of New Jersey~\cite{SMLNJ} and
\ocaml{}~\cite{ocaml-manual-3.0} have used a compilation strategy that optimizes representations of data types when
polymorphic code is monomorphically used, but defaults to a less efficient style otherwise, in order to share
code~\cite{ocaml-repr}.
This strategy reduces code size, but leads to unpredictable performance and code, as changes to a codebase that
either add a new instantiation of a polymorphic function at a given type or, in a modular compilation setting, that
expose a polymorphic function externally, can change the performance of code that is not local to the change being
made.
Monomorphization, as in MLton~\cite{mlton-compiler}, instead instantiates each polymorphic code block at each of the types
it is applied against, providing predictable output code to developers at the cost of some code duplication.
This approach, combined with whole-program compilation~\cite{weeks:whole-program-mlton}, is used in Rust.
This strategy has a negative impact on compilation speed but has resulted in Rust code that easily matches the
sequential speed of its \Cplusplus{} analog without requiring the Servo developers to become compiler experts.

@@

Additionally, Rust's ownership model assumes that there is a single owner for each piece of data.
However, many data structures do not follow that model, in order to provide multiple traversal
APIs without favoring the performance of one over the other.
For example, a doubly-linked list contains a back pointer to the previous element to aid
in traversals in the opposite direction.
Many optimized hashtable implementations also have both hash-based access to items
and a linked list of all of the keys or values.
In Servo, we have had to use unsafe code to implement data structures with this form,
though we are typically able to provide a safe interface to users.

@@

There are two limitations in the language interoperability that pose challenges for Servo today.
First, Rust cannot currently expose varargs-style functions to C code.
Second, Rust cannot compile against \Cplusplus{} code.
In both cases, Servo uses C wrapper code to call into the code that Rust cannot directly
reach.
While this approach is not a large problem for varargs-style functions, it defeats many of the
places where the \Cplusplus{} code has been crafted to ensure the code is inlined into the caller,
resulting in degraded performance.

@@

Compiler plugins can also provide ``lint'' checks\footnote{\url{http://doc.rust-lang.org/book/plugins.html#lint-plugins}} that use the same infrastructure as the compiler's built-in warnings. This allows project-specific safety or style checks to integrate deeply with the compiler. Lint plugins traverse a typechecked abstract syntax tree (AST), and they can be enabled/disabled within any lexical scope, the same way as built-in warnings.
